Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2023-01-05 manuscript.tex
:pencil: f-b ts clust intro full draft

Feature-based time series clustering broadly speaking means that important features from the time series data are extracted (Ratanamahatana & Keogh, 2004). The extracted features can then be used as inputs to a clustering algorithm such as k-means or hierarchical clustering to group the time series into clusters (Xu et al., 2013).

Feature-based time series clustering has been a valued tool in other empirical fields, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}.

As an illustration of the flexibility of this approach, a commonly used software package for this approach allows users to extract a total of 794 time series features \citep[][]{christ2018}.

@@

in psychology:

\citet{kuppens2017} originally proposed (1) variability, (2) co-variance or ICC, (3) inertia, and (4) cross-lagged correlations. These features were then extended and formalized by \citet{krone2018} into (1) within-person variance, (2) innovation, (3) inertia or autocorrelation, (4) cross-lagged regression parameters, (5) covariances as granularity, and (6) mean intensity. \citet{krone2018} were even able to use these features to cluster study participants.

@@

Outside the psychological literature, time series features have also included more summaries in terms of measurement distribution, categorical states, and, importantly, nonlinear developments.

\citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data.

\citet{wang2006} propose a set of nine statistical features for describing a trajectory: (1) trend, (2) seasonality, (3) periodicity, (4) skewness, (5) kurtosis, (6) serial correlation, (7) non-linearity, (8) self-similarity, and (9) chaos. This selection was also used by \citep{fulcher2013}. \citet{adya2001} proposed a total of 28 features relevant for forecasting, which largely fit within the selection proposed by \citet{wang2006}.

@@

\paragraph{Feature extraction}
The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. There are, however, hundreds of possible features to describe psychological time series \citep[e.g., tsfresh][]{christ2018}. One approach to choosing relevant features would be to extract a large number of features and then assess which features are most effective at capturing differences in the time series. However, such an approach is not always advisable for psychological time series. For one, features should reduce the data dimensionality --- it would thus not necessarily be advisable to describe 60 ESM measurements with 180 time series features. More importantly, however, careful feature extraction can be crucial in the interpretability and explainability of the results. This is particularly the case when features have a conceptual or psychological meaning. Taking the concept of well-being as an example, psychologists might be interested in whether certain participants tend to have higher well-being in general (i.e., mean) or whether some participants fluctuate between extremely high and low well-being (i.e., variance). But psychologists might not necessarily be interested in the exact time point after which 50\% of the summed well-being values lie (i.e., relative mass quantile index) or how much different sine wave patterns within the well-being data correlate with one another (i.e., cross power spectral density). We would, thus, strongly advocate for a careful selection of time series features that are meaningful to the field and concepts.

>>>>

The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. In its most general approach, this can include any statistical summaries of the time series \citep[i.e., features; e.g.,][]{maharaj2019}. The extracted features can then be used as inputs to most clustering algorithms, including k-means or hierarchical clustering to group the participants into clusters \citep{wang2006}. The clustering of model parameters is, thus, technically a restrictive sub-form of feature-based clustering --- where lagged regression coefficients, for example, can be considered to be one type of feature capturing a time series. However, beyond model parameters, time series might also be described with any other statistical feature \citep{tiano2021}. Feature-based time series clustering has been a valued tool in many empirical fields for its flexibility and interpretability. As a result, features have been extracted for a variety of data, including analyses of astronomical, meteorological, and aviation pathways, biological and medical developments, as well as energy and finance patterns \citep{Aghabozorgi2015}.

With its rich history, a staggering variety of time series features have been proposed, not all of which are meaningful for psychological time series. As an example, \citet{wang2006} propose a set of nine statistical features for describing a trajectory: (1) trend, (2) seasonality, (3) periodicity, (4) skewness, (5) kurtosis, (6) serial correlation, (7) non-linearity, (8) self-similarity, and (9) chaos. This selection was also used by \citep{fulcher2013}. \citet{adya2001} proposed a total of 28 features relevant for forecasting, which largely fit within the selection proposed by \citet{wang2006}. A final illustration comes from a commonly used software package for feature extraction `tsfresh', which allows users to extract a total of 794 features to describe all kinds of aspects of a time series \citep[][]{christ2018}.

Given that there are hundreds of possible features to describe psychological time series, one approach to choosing relevant features would be to extract a large number of features and then assess which features are most effective at capturing differences in the time series \citep[e.g.,][]{christ2018}. However, such an approach is not always advisable for psychological time series. For one, features should reduce the data dimensionality --- it would thus not necessarily be advisable to describe 60 ESM measurements with 794 time series features. More importantly, however, careful feature extraction can be crucial in the interpretability and explainability of the results. This is particularly the case when features have a conceptual or psychological meaning. Taking the concept of well-being as an example, psychologists might be interested in whether certain participants tend to have higher well-being in general (i.e., mean) or whether some participants fluctuate between extremely high and low well-being (i.e., variance). But psychologists might not necessarily be interested in the exact time point after which 50\% of the summed well-being values lie (i.e., relative mass quantile index) or how much different sine wave patterns within the well-being data correlate with one another (i.e., cross power spectral density). We will, thus, propose and describe a set of time series features that have a strong backing within the ESM literature and as such offer a reasonable starting point for most concepts within the field.

@@

\citet{kuppens2017} originally proposed (1) variability, (2) co-variance or ICC, (3) inertia, and (4) cross-lagged correlations. These features were then extended and formalized by \citet{krone2018} into (1) within-person variance, (2) innovation, (3) inertia or autocorrelation, (4) cross-lagged regression parameters, (5) covariances as granularity, and (6) mean intensity. \citet{krone2018} even built a parametric model to use a subset of these features to cluster study participants.

@@

Time series features have also come to include more summaries in terms of measurement distribution, categorical states, and, importantly, nonlinear developments. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data.

@@

\paragraph{Feature extraction}
more technical description model + descriptive stats
% The main aim of feature extraction is to describe the most important and meaningful aspects of a time series. There are, however, hundreds of possible features to describe psychological time series \citep[e.g., tsfresh][]{christ2018}. One approach to choosing relevant features would be to extract a large number of features and then assess which features are most effective at capturing differences in the time series. However, such an approach is not always advisable for psychological time series. For one, features should reduce the data dimensionality --- it would thus not necessarily be advisable to describe 60 ESM measurements with 180 time series features. More importantly, however, careful feature extraction can be crucial in the interpretability and explainability of the results. This is particularly the case when features have a conceptual or psychological meaning. Taking the concept of well-being as an example, psychologists might be interested in whether certain participants tend to have higher well-being in general (i.e., mean) or whether some participants fluctuate between extremely high and low well-being (i.e., variance). But psychologists might not necessarily be interested in the exact time point after which 50\% of the summed well-being values lie (i.e., relative mass quantile index) or how much different sine wave patterns within the well-being data correlate with one another (i.e., cross power spectral density). We would, thus, strongly advocate for a careful selection of time series features that are meaningful to the field and concepts.
